---
title: "Quick Start"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7,
  fig.height = 5
)
library(framedf)
```

## What `framedf` Does

`framedf` is a first-pass diagnostic for an unfamiliar data frame. It tries
to read the data the way a careful colleague would — by figuring out what
each column means, by noticing pairs of columns that move together, and by
flagging values that look wrong.

The result is qualitative. You get sentences like *"temperature decreases
strongly with elevation"*, not Pearson coefficients. Numeric evidence is
still on the object for anyone who wants it; the printed view stays calm.

There is one entry point, `frame()`, and four reader functions:

* `print(frame(df))` — narrative overview.

* `relationships(frame(df))` — pairs grouped as meaningful, suspicious,
  structural, or ignored.
* `anomalies(frame(df))` — per-column oddities.

* `details(frame(df))` — methods, roles, skipped rules, backend.

## Installation

```{r, eval = FALSE}
# install.packages("pak")
pak::pak("gcol33/framedf")
```

## A First Look

```{r}
set.seed(1)
n <- 300L
df <- data.frame(
  plot_id    = sample(1:60, n, replace = TRUE),
  year       = sample(2010:2020, n, replace = TRUE),
  latitude   = stats::runif(n, 40, 50),
  longitude  = stats::runif(n, 5, 15),
  elevation  = stats::runif(n, 0, 2500),
  observer   = sample(letters[1:6], n, replace = TRUE),
  country    = sample(c("AT", "DE", "CH"), n, replace = TRUE),
  richness   = NA_real_,
  cover_a    = stats::runif(n, 0, 1),
  stringsAsFactors = FALSE
)
df$temperature <- 20 - df$elevation / 200 + stats::rnorm(n)
df$richness    <- 5 + 0.4 * sqrt(stats::runif(n, 1, 100)) +
                  2 * (df$observer == "a") + stats::rnorm(n)
df$cover_b     <- 1 - df$cover_a
```

```{r}
fd <- frame(df)
print(fd)
```

The four sections of the print output map onto the four parts of any
first-look conversation:

* **Structure** — what kind of dataframe is this? How is it shaped, what
  identifies a row, what is the temporal/spatial axis?
* **Relationships** — what moves with what? Which patterns are real and
  which feel like artefacts?
* **Anomalies** — what looks wrong?

* **Ignored** — what was excluded from screening, and why?

## Drill Down

```{r}
relationships(fd)
```

```{r}
anomalies(fd)
```

```{r}
details(fd)
```

`details()` is the place to go when an output surprises you. It tells you
which mode the screening ran in, what role each column was given, which
rules caused some pairs to be skipped, and whether the C++ backend was
used.

## Adjust for a Confounder

If you suspect one variable is confounding most of the others, you can
partial it out before screening:

```{r}
fd_adj <- frame(df, adjustment = "elevation")
relationships(fd_adj, kind = "meaningful")
```

## Tune the Strength Tiers

Everything is configurable through `framedf_settings()` or the `...` of
`frame()`:

```{r, eval = FALSE}
frame(df,
      strong_threshold   = 0.6,
      moderate_threshold = 0.4,
      weak_threshold     = 0.15)
```

## Next Steps

* [Workflows](workflows.html) — practical examples (large data, repeated
  measures, observer effects).
* [Roles and rules](roles.html) — what each role means and why some
  columns are excluded from screening.
