---
title: "Roles and Rules"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Roles and Rules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(framedf)
```

`framedf` decides what to do with a column based on the **role** it
infers. This vignette documents every role, the cues used to assign it,
and the screening behaviour it triggers.

## Why Roles Matter

A correlation matrix treats `row_id` and `temperature` identically. They
are both numeric, after all. But conceptually they are not the same
column — `row_id` is bookkeeping, and a strong correlation between
`row_id` and `temperature` is almost always a sign that the data is
sorted by time, not a real association.

Role inference is what lets `framedf` say *"latitude was ignored
because it is a spatial coordinate, screened separately as drift"*
instead of just dumping a Pearson coefficient.

## The Roles

| Role            | Cues used to assign it                                                                       | Screening behaviour                                                                  |
|-----------------|-----------------------------------------------------------------------------------------------|--------------------------------------------------------------------------------------|
| `id`            | All-unique integer with id-like name; or character with very high uniqueness ratio.          | Excluded from the numeric pair sweep. Listed in **Ignored**.                          |
| `admin_index`   | Sequential integer with a row-number-like name.                                               | Excluded.                                                                            |
| `group_id`      | Repeated integer or repeated character with id-like name.                                     | Excluded from the symmetric numeric sweep, but used as a *grouping* in cat–num scan. |
| `temporal`      | Date or POSIX class; or integer in 1500–2200 with a year-like name.                           | Excluded from the symmetric sweep; participates in the drift screen.                 |
| `coord_lat`     | Name matches `lat`/`latitude` and the bulk of values fall within `[-90, 90]`.                 | Excluded; participates in the drift screen.                                          |
| `coord_lon`     | Name matches `lon`/`lng`/`longitude` and the bulk of values fall within `[-180, 180]`.        | Excluded; participates in the drift screen.                                          |
| `measurement`   | Continuous numeric without any other cue.                                                     | Screened symmetrically against all other measurements (and compositionals).          |
| `compositional` | Numeric with a name like `*cover*`, `*pct*`, `*share*`, `*proportion*`, in `[0, 1]` or `[0, 100]`. | Screened, plus checked pairwise for constrained-complement structure.            |
| `categorical`   | Character or factor with low cardinality.                                                     | Used as a grouping in the cat–num scan.                                              |
| `flag`          | Logical.                                                                                      | Treated like a categorical with two levels.                                          |
| `near_constant` | Dominant value covers more than `(1 - near_constant_ratio)` of rows.                          | Excluded.                                                                            |
| `constant`      | Only one unique value.                                                                        | Excluded.                                                                            |
| `unknown`       | None of the above.                                                                            | Excluded.                                                                            |

## Worked Example

```{r}
df <- data.frame(
  PlotID            = sample(1:50, 200, replace = TRUE),
  PlotObservationID = seq_len(200L),
  row_number        = seq_len(200L),
  year              = sample(1990:2020, 200, replace = TRUE),
  latitude          = stats::runif(200, 40, 50),
  longitude         = stats::runif(200, 5, 15),
  elevation         = stats::runif(200, 0, 2500),
  observer_id       = sample(letters[1:5], 200, replace = TRUE),
  country           = sample(c("AT", "DE", "CH"), 200, replace = TRUE),
  native_cover      = stats::runif(200, 0, 1),
  flagged           = sample(c(TRUE, FALSE), 200, replace = TRUE),
  stringsAsFactors  = FALSE
)
df$alien_cover <- 1 - df$native_cover
fd <- frame(df)
details(fd)
```

The **Column roles** block in `details()` lists every column with its
inferred role. Anything in the **Ignored** section of `print(fd)` will
also be there.

## Tuning the Rules

Every threshold is a parameter:

```{r, eval = FALSE}
frame(df,
      id_unique_ratio     = 0.99,   # require 99% unique to call a string an id
      near_constant_ratio = 0.01,   # call near-constant if dominant > 99%
      strong_threshold    = 0.6,
      moderate_threshold  = 0.4,
      weak_threshold      = 0.15,
      compositional_cv    = 0.05,
      observer_min_levels = 4L,
      min_level_n         = 5L)
```

`framedf_settings()` returns the full default list with documentation
for every field.

## When the Role Is Wrong

Role inference is heuristic, and edge cases will land in the wrong
bucket sometimes. Two ways to recover:

1. Rename the column. Names like `*_id`, `*_year`, `*_lat`, `*_cover`
   are picked up immediately.
2. Read the role from `fd$roles` and override the analysis manually:

```{r, eval = FALSE}
fd <- frame(df)
fd$roles[["my_weird_column"]] <- "measurement"
# (re-run any reader you need on the modified object)
```

The triage object is a plain list, so post-hoc edits are safe. For
production use, prefer fixing the column name or extending the
package's name-based heuristics.
